---
title: "FinalProj_Measles_pres"
author: "Rory Sauereisen"
format: revealjs
editor: visual
---

```{r}
#| include: false
#| message: false
#| warning: false
##load in data and libraries here w output == false, so not included in slides
#Reproducible -so it needs library and data load-ins 
#for presentations, increase axis sizes! (will look odd in window, but great on slides)
# theme_minimal(base_size = 24)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(plotly)
library(here)
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
```

## Data Introduction

-   TidyTuesday: "Measles cases across the world"
-   Measles cases rising

```{r}
(cases_year |>
  filter(country == "United States of America") |>
  filter(year == 2025 | year == 2024) |>
  select(country, year, total_population, measles_total, measles_incidence_rate_per_1000000_total_population, rubella_total, rubella_incidence_rate_per_1000000_total_population))
```

## Question(s) of Interest
- What are the top 10 countries with the highest record cases of Measles? How does the USA compare to these countries?
-   ^ Throughout the data's timeline of measles counts, is there any evidence of historical impacts? 
- Is there a visible global rise in Measles cases since 2012?


## Top 10 Countries With the Highest Measles Counts
```{r}
#| message: false
#| warning: false
#| include: false
top10_yearspan_measles <-
  cases_year |>
  group_by(country) |>
  arrange(desc(measles_total)) |>
  slice(1) |>
  ungroup(country) |>
  arrange(desc(measles_total)) |>
  slice(1:10) |>  # Now take the names of just those top countries:
  distinct(country) |> 
  pull(country)

# Filter the dataset to only include the top historical measles countries, and the USA for comparison
top10nUSA_measles_tot <-
  cases_year |>
  filter(country %in% top10_yearspan_measles | country == "United States of America") |>
  mutate(hand_incidenceR_measles = (measles_total / total_population)*1000000) #updated incidence rate
```
```{r}
#Looking @ counts:
top10measlesN_plot <-
ggplot(data = top10nUSA_measles_tot, aes(x = year, y = measles_total)) +
  geom_line(aes(color = country), linewidth = 1) +
  theme_minimal()


ggplotly(top10measlesN_plot, tooltip = c("color", "x", "y"))
```

# These Top 10 Countries: Measles Incidence rates
```{r}
#Looking @ Incidence rate:
top10measlesI_plot <-
ggplot(data = top10nUSA_measles_tot, aes(x = year, y = hand_incidenceR_measles)) +
  #facet_wrap(~region) +
  geom_line(aes(color = country), linewidth = 1) +
  theme_minimal() +
  labs(y = "Measles cases per 1,000,000")


ggplotly(top10measlesI_plot, tooltip = c("color", "x", "y"))
```


# Over Time: Global Increase? 
```{r}
#| message: false
#| warning: false
#| include: false
world_map <- map_data("world")
cases_2012year <- #filter to start, fix country names to show up on map
  cases_year |>
  filter(year == 2012) |>
  mutate(country = if_else(country == "United States of America", 
                           true = "USA", 
                           false = country)) |>
  mutate(country = if_else(country == "Russian Federation", 
                           true = "Russia", 
                           false = country)) |>
  mutate(country = if_else(country == "Venezuela (Bolivarian Republic of)", 
                           true = "Venezuela", 
                           false = country)) |>
  mutate(country = if_else(country == "Iran (Islamic Republic of)", 
                           true = "Iran", 
                           false = country)) |>
  mutate(country = if_else(country == "Democratic People's Republic of Korea", 
                           true = "North Korea", 
                           false = country)) |>
  mutate(country = if_else(country == "Bolivia (Plurinational State of)", 
                           true = "Bolivia", 
                           false = country)) |>
  mutate(country = if_else(country == "United Kingdom of Great Britain and Northern Ireland", 
                           true = "UK", 
                           false = country)) |>
  mutate(country = if_else(country == "Netherlands (Kingdom of the)", 
                           true = "Netherlands", 
                           false = country)) 

cases_year |>
  filter(year == 2012) |>
  filter(country == "United States of America")

#join the datasets:
full_worldMeasles <- left_join(world_map, cases_2012year, 
                           join_by(region == country))
```
```{r}
#plot the map:
ggplot(data = full_worldMeasles, aes(x = long, y = lat, 
                                 group = group)) +
  geom_polygon(aes(fill=measles_total)) +
  scale_fill_viridis_c() +
  coord_map(projection = "mercator") +
  xlim(c(-180,180)) +
  labs(title = "Cases of Measles in 2012")
```

# Over Time: Global Increase? 
```{r}
#| message: false
#| warning: false
#| include: false
world_map <- map_data("world")
cases_2018year <- #filter to start, fix country names to show up on map
  cases_year |>
  filter(year == 2018) |>
  mutate(country = if_else(country == "United States of America", 
                           true = "USA", 
                           false = country)) |>
  mutate(country = if_else(country == "Russian Federation", 
                           true = "Russia", 
                           false = country)) |>
  mutate(country = if_else(country == "Venezuela (Bolivarian Republic of)", 
                           true = "Venezuela", 
                           false = country)) |>
  mutate(country = if_else(country == "Iran (Islamic Republic of)", 
                           true = "Iran", 
                           false = country)) |>
  mutate(country = if_else(country == "Democratic People's Republic of Korea", 
                           true = "North Korea", 
                           false = country)) |>
  mutate(country = if_else(country == "Bolivia (Plurinational State of)", 
                           true = "Bolivia", 
                           false = country)) |>
  mutate(country = if_else(country == "United Kingdom of Great Britain and Northern Ireland", 
                           true = "UK", 
                           false = country)) |>
  mutate(country = if_else(country == "Netherlands (Kingdom of the)", 
                           true = "Netherlands", 
                           false = country)) 


#join the datasets:
full_worldMeasles <- left_join(world_map, cases_2018year, 
                           join_by(region == country))
```
```{r}
#plot the map:
ggplot(data = full_worldMeasles, aes(x = long, y = lat, 
                                 group = group)) +
  geom_polygon(aes(fill=measles_total)) +
  scale_fill_viridis_c() +
  coord_map(projection = "mercator") +
  xlim(c(-180,180)) +
  labs(title = "Cases of Measles in 2018")
```



# Over Time: Global Increase? 
```{r}
#| message: false
#| warning: false
#| include: false
world_map <- map_data("world")
cases_2024year <- #filter to start, fix country names to show up on map
  cases_year |>
  filter(year == 2024) |>
  mutate(country = if_else(country == "United States of America", 
                           true = "USA", 
                           false = country)) |>
  mutate(country = if_else(country == "Russian Federation", 
                           true = "Russia", 
                           false = country)) |>
  mutate(country = if_else(country == "Venezuela (Bolivarian Republic of)", 
                           true = "Venezuela", 
                           false = country)) |>
  mutate(country = if_else(country == "Iran (Islamic Republic of)", 
                           true = "Iran", 
                           false = country)) |>
  mutate(country = if_else(country == "Democratic People's Republic of Korea", 
                           true = "North Korea", 
                           false = country)) |>
  mutate(country = if_else(country == "Bolivia (Plurinational State of)", 
                           true = "Bolivia", 
                           false = country)) |>
  mutate(country = if_else(country == "United Kingdom of Great Britain and Northern Ireland", 
                           true = "UK", 
                           false = country)) |>
  mutate(country = if_else(country == "Netherlands (Kingdom of the)", 
                           true = "Netherlands", 
                           false = country)) 


#join the datasets:
full_worldMeasles <- left_join(world_map, cases_2024year, 
                           join_by(region == country))
```
```{r}
#plot the map:
ggplot(data = full_worldMeasles, aes(x = long, y = lat, 
                                 group = group)) +
  geom_polygon(aes(fill=measles_total)) +
  scale_fill_viridis_c() +
  coord_map(projection = "mercator") +
  xlim(c(-180,180)) +
  labs(title = "Cases of Measles in 2024")
```


