---
title: "Measles Cases - Final Project"
author: "Aurora Sauereisen"
format: 
  html:
    embed-resources: true
---
The purpose of this final project is to explore global measles cases, and showcase data story telling abilities

Data Chosen from TidyTuesday GitHub. 
Title & Post date: 'Measles cases across the world', 2025-06-24
Link: (https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-06-24)
Data is measel case counts from 2012 to 2025

*Read in the data:*
```{r}
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')


#Having issues saving said data to my computer itself. 
write_csv(cases_month, "C:\Users\sauer\OneDrive\Documents\St. Lawrence University\2026\MeaslesCases_334proj\data")

```

*Purpose: Visualize by country and timeline the severity and spread of measles*
Questions to explore:
 - For each region, what season had the highest # measels cases? 
 - Are we able to reveal differences in healthcare quality between regions/countries, based on observing their ratios of laboratory to suspected- confirmed cases of disease? (Does the ratio of laboratory-confirmed cases to total cases reveal differences in healthcare capacity across countries?)
 - Is there a trend in measles vs rubella prominence? Does this show up regionally or by country, suggesting environmental or public health influencers?


 
 
 
# Exploring Measles over the Years
Line plot: #measles vs year, facet by region, line by country. Leaflet plot to show # cases at that time point based on cursor hover
Second plot: #measles should be the incidence rate (so country populations in comparison can be considered) Then talk about burden of disease on countries. Like even though they have small numbers, that's a large portion of their population, bringing to light the burden/impact on the country. 

To simplify, looking at top 15 countries with the highest rates across the years.  
```{r}
#From all the years, take each country's highest count.
# Then only take the the top 10 countries with their record high counts
library(tidyverse)
library(dplyr)

#Top 10 countries that have had the record highest total measles cases during the span of 2012 - 2025:
top10_yearspan_measles <-
  cases_year |>
  group_by(country) |>
  arrange(desc(measles_total)) |>
  slice(1) |>
  ungroup(country) |>
  arrange(desc(measles_total)) |>
  slice(1:10) |>
  distinct(country) |> # Now take the names of just those top countries
  pull(country)

# Filter the dataset to only include the top historical measles countries, and the USA for comparison
top10nUSA_measles_tot <-
  cases_year |>
  filter(country %in% top10_yearspan_measles | country == "United States of America")
```

Plot:
```{r}
library(ggplot2)
library(plotly)
top10measlesN_plot <-
ggplot(data = top10nUSA_measles_tot, aes(x = year, y = measles_total)) +
  #facet_wrap(~region) +
  geom_line(aes(color = country), linewidth = 1) +
  theme_minimal()

ggplotly(top10measlesN_plot, tooltip = c("color", "x", "y"))
# ^can't get it to order the way I'd like?


#Looking at Incidence rate:
top10measlesI_plot <-
ggplot(data = top10nUSA_measles_tot, aes(x = year, y = measles_incidence_rate_per_1000000_total_population)) +
  #facet_wrap(~region) +
  geom_line(aes(color = country), linewidth = 1) +
  theme_minimal()

ggplotly(top10measlesI_plot, tooltip = c("color", "x", "y"))
```
Between these two plots, we see some differences between Measeles count versus the incidence rate of the disease in eachcountry, which gives insight as to the effect of measles presence on the country hollistically. *PLEASE CONTINUE HERE*


# Measles vs Rubella: Trends in co-existence(?)
In the top countries with the highest measles prevalence, how does their rubella count compare to their measles count? 
