---
title: "Measles Cases - Final Project"
author: "Aurora Sauereisen"
format: 
  html:
    embed-resources: true
---
The purpose of this final project is to explore a dataset of choice/interest and showcase data story telling abilities.

# Introduction
It begins with a rash on the face, spreading to the chest and back before the rest of the body, combined with fever (up to 105F), a dry cough, runny nose, and conjunctivitis. Highly contagious through airborne and physical contact routes, people of any age can contract Measles. While children are more often exposed to & contract Measles, in most cases a patient is bedridden for 7-10 days. For the immune compromised and those with complications though, this can be a life threatening disease and needs medical monitoring to limit life-threatening complications. However, with the first vaccines developed and licensed in 1963, the U.S. all but declared the Measles virus eliminated in the U.S. by the year 2000. \
\
But as current media campaigns warn, Measles has all but renewed it's presence in the U.S.; -with alarming foothold. This blog entry aims to explore a data set sourced from the GitHub repository 'TidyTuesday' covering global Measles case data. Exploring global performance in addition to examining U.S. specific rates, this blog will aim to visualize Measles's presence in the world. \
**Blog Goals:**
- What are the top 10 countries with the highest record cases of Measles? How does the USA compare to these countries?
- Throughout the data's timeline of measles counts, is there any evidence of historical impacts? 
- Are we able to reveal differences in healthcare quality between regions/countries, based on observing their ratios of laboratory to suspected- confirmed cases of disease? (Does the ratio of laboratory-confirmed cases to total cases reveal differences in healthcare capacity across countries?)
 - Is there a trend in measles vs rubella prominence? Does this show up regionally or by country, suggesting environmental or public health influencers?


*Read in the data:*
```{r}
#| message: false
#| warning: false
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
```

*Purpose: Visualize by country and timeline the severity and spread of measles*
Questions to explore:
 - For each region, what season had the highest # measles cases? 
 - Are we able to reveal differences in healthcare quality between regions/countries, based on observing their ratios of laboratory to suspected- confirmed cases of disease? (Does the ratio of laboratory-confirmed cases to total cases reveal differences in healthcare capacity across countries?)
 - Is there a trend in measles vs rubella prominence? Does this show up regionally or by country, suggesting environmental or public health influencers?


 
# Exploring Measles over the Years
Line plot: #measles vs year, facet by region, line by country. Leaflet plot to show # cases at that time point based on cursor hover
Second plot: #measles should be the incidence rate (so country populations in comparison can be considered) Then talk about burden of disease on countries. Like even though they have small numbers, that's a large portion of their population, bringing to light the burden/impact on the country. 

To simplify, looking at top 15 countries with the highest rates across the years:  
```{r}
library(tidyverse)
library(dplyr)

#Top 10 countries that have had the record highest total measles cases during the span of 2012 - 2025:
top10_yearspan_measles <-
  cases_year |>
  group_by(country) |>
  arrange(desc(measles_total)) |>
  slice(1) |>
  ungroup(country) |>
  arrange(desc(measles_total)) |>
  slice(1:10) |>  # Now take the names of just those top countries:
  distinct(country) |> 
  pull(country)

# Filter the dataset to only include the top historical measles countries, and the USA for comparison
top10nUSA_measles_tot <-
  cases_year |>
  filter(country %in% top10_yearspan_measles | country == "United States of America") |>
  mutate(hand_incidenceR_measles = (measles_total / total_population)*1000000) #updated incidence rate
```

Plot:
```{r}
library(ggplot2)
library(plotly)
top10measlesN_plot <-
ggplot(data = top10nUSA_measles_tot, aes(x = year, y = measles_total)) +
  #facet_wrap(~region) +
  geom_line(aes(color = country), linewidth = 1) +
  theme_minimal()

ggplotly(top10measlesN_plot, tooltip = c("color", "x", "y"))
# ^can't get it to order the way I'd like?


#Looking at Incidence rate:
top10measlesI_plot <-
ggplot(data = top10nUSA_measles_tot, aes(x = year, y = hand_incidenceR_measles)) +
  #facet_wrap(~region) +
  geom_line(aes(color = country), linewidth = 1) +
  theme_minimal() +
  labs(y = "Measles cases per 1,000,000")

ggplotly(top10measlesI_plot, tooltip = c("color", "x", "y"))
```
Between these two plots, we see some differences between Measles count versus the incidence rate of the disease in each country. The *incidence rate* of Measles in the second graph displays the country's case count while accounting for the overall population. While the incidence rate visualization gives a better understanding as to how the disease may be impacting the country overall, it also seems to water down or downplay the fluctuation and presence of Measles.  \
Madagascar in particular seems to be heavily affected by Measles, meanwhile many other countries trail and fluctuate at case counts about half of Madagascar's. 
*PLEASE CONTINUE HERE*

Something else to note is the collective drop in measles cases in all countries shortly following the year 2020. This is likely due to the Covid-19 pandemic, where social distancing and extra sanitation likely limited the spread of measles in addition to Covid-19 during that time period. However, another possible reason behind the drop in cases could be from under reporting; as Covid-19 was at the forefront of health professionals and the media's attention, measles cases could have been triaged in data collection or not brought in for medical diagnosis. As I do not have complementary data on hand or expertise to discern which (or neither) of these ideas are responsible for this observation, I will not report in confidence a reason for this trend. \

# Measles & the U.S.:
In comparison to the top 10 countries with the highest counts of measles, the United States seems to pale in comparison. But the reason this data set was posted on GitHub's TidyTuesday repository was to mirror the awareness campaigns of the rising rates of measles in the United states. A quick glimpse of the US's measles counts and incidence rate captures this concerning return to rising measles rates after Covid. \

```{r}
cases_year_USA <- 
  cases_year |>
  filter(country == "United States of America") |>
  mutate(hand_incidenceR_measles = (measles_total / total_population)*1000000) |>
  select(hand_incidenceR_measles, measles_incidence_rate_per_1000000_total_population, year, 
         annualized_population_most_recent_year_only, total_population, everything())
  
  
measlesUSA_plot <-
ggplot(data = cases_year_USA, aes(x = year, y = measles_incidence_rate_per_1000000_total_population, label1 = total_population, label2 = measles_total)) +
  geom_line(linewidth = 1) +
  geom_point(aes(x = year, y = hand_incidenceR_measles), color = "purple") +
  theme_minimal()

ggplotly(measlesUSA_plot, tooltip = c("y", "label1", "label2"))
```
*A critical adjustment here:* While investigating US incidence rates, I identified that there's an unusual value for 2025 that gives a false interpretation of a drastic increase in cases. While my hand calculation for the 2025 incidence rate reveals a rate of approximately 3.4 cases of measles per 100,000 people, the dataset reports a value of approximately 8 cases of measles per 100,000 people. My hand calculation results mirror the rest of the values in the data set except for this one 2025 value. \
    This is reflected in the graph as the black line with data-original values and purple plotted points to represent my hand calculations to check it. The difference is due to the original data using the annualized population opposed to total population for only the year 2025. To limit misinterpretations and maintain consistency across reported values, all visualizations in this report will use my hand-calculated incidence rate values (the purple points) that not utilize the country-wide 2025 annualized population values. (The above section examining global counts incorporates this same adjustment to relying on my personal calculations) \
  \
Regardless of the 2025 incidence rate value, this plotly line plot delivers the same message; the return in Measles cases after the 2020 pandemic like the rest of the globe. However, just how bad is measles getting in the US? \
\
The data set currently visualized here is from a github repository dating from April 24, 2025, about a year ago from now, and only reports on country-level counts. While it warns of an increasing trend after Covid-19, it doesn't capture the increasing concern since that date, and at a global level at that. I've pulled here supplementary data from the CDC to get a glimpse at the more current case counts and rates of measles cases to gain a better understanding of the U.S.'s current standing with the disease. \
  \
# Current Glimpse: 2025-26 U.S. Measles Cases
..> also show vaccine counts? 
..> do a map of USA states with diverging color scale based on an increase in counts or decrease, and by how much from the previous year. Also a month by month line plot? Maybe separate lines by region of the US? 


# Measles vs Rubella: Trends in co-existence(?)
While measles are known for ___ symptoms, rubella compares in the sense of ____. 
In the top countries with the highest measles prevalence, how does their rubella count compare to their measles count? The gitHub data set includes rubella counts in addition to measles. Taking advantage of the extra information, lets visualize: \
Pick a year and boxplots? 
```{r}
#pivot the data
cases_year |>
  select(region, country, year, total_population, measles_total, rubella_total) |>
  pivot_longer(cols = c(measles_total, rubella_total),
             names_to = "disease", values_to = "total_cases") |>
  select(total_cases, disease, everything()) |>
  #finish this
```



# Conclusions

# Notes:
-> **Incorrect Incidence rate data**. While investigating US measles incidence rate over time (`measles_incidence_rate_per_1000000_total_population` variable), I discovered that the 2025 value was incorrect. In hand calculating the yearly incidence rates for the US, all were correct except the 2025 value. Unsure of how the data authors calculated their values to identify the calculation that produced it, I chose not to use the current column for measles incidence rate in the 'case_year' data set.  
- If given more time, I would investigate Madigascar's high rates of Measles more thoroughly. 

## Sources:
**Main data set explored:** \
'Measles cases across the world', posted 2025-06-24  \
Chosen from TidyTuesday GitHub repository. Contains information on global Measles and Rubella cases.  \
Data set temporal range: 2012 - 2025 \
Link: (https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-06-24)  \
 \
 \
**Measles Supplementary Information:** \
- Mayo Clinic 'Measles' - Symptoms & Causes page
   Link: (https://www.mayoclinic.org/diseases-conditions/measles/symptoms-causes/syc-20374857)
- Mayo Clinic 'History of infectious disease outbreaks and vaccines timeline' - Measles page
   Link: (https://www.mayoclinic.org/diseases-conditions/history-disease-outbreaks-vaccine-timeline/measles)
